---
title: "Geometric typicality: testing an effect in a principal plane"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Geometric typicality: testing an effect in a principal plane}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
needed <- c("carData", "FactoMineR", "ggplot2")
ok <- all(vapply(needed, requireNamespace, logical(1), quietly = TRUE))
knitr::opts_chunk$set(
  collapse = TRUE, comment = "#>", eval = ok,
  fig.width = 6.5, fig.height = 5, fig.align = "center"
)
```

## The question

When the same individuals are observed in two conditions -- *before* and *after*
a treatment -- each one carries an **effect**: the deviation from its "before"
point to its "after" point. If the observations are multidimensional these
effects are *vectors*, and they live in the geometric space of a GDA. The
natural question is whether the **mean effect-vector** is a genuine one, or could
just be the luck of the draw.

This is the situation the **geometric typicality test** is made for (Le Roux,
Bienaise & Durand 2019, ch. 4; their Parkinson case study, §6.1). The reasoning
is a *reflection* argument:

> If the treatment did nothing, the "before" and "after" measurements of a
> subject would be interchangeable -- so each subject's effect-vector would be
> just as likely to point one way as the opposite. Flipping the sign of each
> effect-vector in every possible way builds the reference distribution, and we
> ask how far the observed mean effect is from the **null point** $O$ ("no
> effect").

`typicality_geom()` does this, in the principal plane of a GDA, returning a
p-value and a compatibility region.

```{r load}
library(GDAinference)
library(FactoMineR)
library(ggplot2)
```

## Building the cloud: a GDA of the response profiles

We use the O'Brien--Kaiser repeated-measures data (`carData`): 16 subjects in a
`control` group and two treatment groups (`A`, `B`), each measured on five
successive occasions, **before** and **after** the intervention.

The five occasions form a *response profile*. We summarise these profiles by a
PCA -- pooling the before and after profiles so that both are placed in the same
space:

```{r pca}
data("OBrienKaiser", package = "carData")
pre  <- as.matrix(OBrienKaiser[, paste0("pre.",  1:5)])
post <- as.matrix(OBrienKaiser[, paste0("post.", 1:5)])

pca <- PCA(rbind(pre, post), scale.unit = TRUE, ncp = 5, graph = FALSE)
round(pca$eig[1:3, "percentage of variance"], 1)
round(pca$var$coord[, 1:2], 2)
```

The first plane is easy to read. **Axis 1** (80% of the variance) has all five
loadings positive: it is an *overall response level* (high on every occasion vs.
low on every occasion). **Axis 2** opposes the early occasions to the late ones:
it is a *within-session shape*. Each subject now has a "before" point and an
"after" point in this plane.

## Effect-vectors

The effect of the intervention for each subject is the vector from its "before"
point to its "after" point -- a difference of principal coordinates:

```{r effects}
co <- get_coord(pca)
effect <- co[17:32, 1:2] - co[1:16, 1:2]      # post - pre, in the first plane
colnames(effect) <- c("axis 1: overall level", "axis 2: shape")
head(round(effect, 2))
```

These 16 effect-vectors form a cloud. Under "no effect", each one is equally
likely to point either way (the reflection argument above), so the geometric
typicality test sign-flips them and asks whether their **mean** is displaced from
$O = (0, 0)$.

## The test

```{r by-group}
treatment <- OBrienKaiser$treatment
do.call(rbind, lapply(levels(treatment), function(g) {
  res <- typicality_geom(effect[treatment == g, ], point = c(0, 0), seed = 1)
  data.frame(group = g, n = res$n,
             D = round(res$statistic, 2),
             p_value = round(res$p_value, 4),
             notable = res$notable)
}))
```

The contrast is clear -- with the caveat that group `A` (n = 4) is far too small
for the test to reach significance (with only $2^{3}$ sign-flips the smallest
attainable p-value is large). Pooling the two treatment groups gives adequate
size:

```{r treated}
res_control <- typicality_geom(effect[treatment == "control", ], point = c(0, 0))
res_treated <- typicality_geom(effect[treatment %in% c("A", "B"), ],
                               point = c(0, 0), seed = 1)
res_treated
```

* The **control** group's mean effect is essentially at $O$ (`D = 0.15`,
  `p = 1`): its before/after measurements are interchangeable -- exactly what
  "no effect" should look like.
* The **treated** group's mean effect is notable and atypical (`D = 2.07`,
  `p = 0.002`): the intervention produced a genuine shift.

The `plot()` shows *where* the treated group moved:

```{r plot, fig.alt = "Effect-vectors of the treated group with the 95% compatibility ellipse; the null point O lies outside it."}
plot(res_treated)
```

The cloud of effect-vectors has its mean **G** well to the right of the null
point **P** = $O$, along **axis 1**: after treatment, responses rose to a higher
*overall level*, with little change of *shape* (axis 2). Because **P** lies
outside the dashed compatibility ellipse, the mean effect is significant at the
5% level. Had the effect been null -- as for the control group -- $O$ would have
fallen inside the region.

## Why *geometric* typicality here?

It is worth contrasting the two typicality tests of the package on this example:

* [typicality_comb()] (combinatorial, ch. 3) treats a group as a *sample* of a
  reference population and permutes which individuals belong to it. That is the
  right test for "is this subgroup atypical of the whole cloud?" (see the
  `vignette("typicality")`).
* `typicality_geom()` (geometric, ch. 4) treats the effect-vectors as a cloud in
  their own right and tests its mean against a fixed *point*. There is no
  reference population to sample from here -- only a within-subject reflection
  symmetry -- so this is the appropriate test.

## Practical notes

* **Exact vs. Monte Carlo.** Up to ~17 subjects the $2^{n-1}$ sign-flips are
  enumerated exactly; beyond that, `max_samples` of them are drawn at random
  (set `seed`).
* **The principal plane.** The test is run on the axes you pass (`axes`); read it
  on the plane you interpret, as in any GDA.
* **The reference point.** `point` defaults to the origin (`0` = "no effect");
  pass any point to test a different null.

## Reference

Le Roux, B., Bienaise, S. & Durand, J.-L. (2019).
*Combinatorial Inference in Geometric Data Analysis*. Chapman & Hall/CRC
(the Parkinson study, §6.1).
