---
title: "Homogeneity: do two groups differ in a GDA cloud?"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Homogeneity: do two groups differ in a GDA cloud?}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
needed <- c("questionr", "FactoMineR", "ggplot2")
ok <- all(vapply(needed, requireNamespace, logical(1), quietly = TRUE))
knitr::opts_chunk$set(
  collapse = TRUE, comment = "#>", eval = ok,
  fig.width = 7, fig.height = 5.2, fig.align = "center"
)
```

## The question

A Geometric Data Analysis lays out groups as mean points in a cloud, and it is
tempting to read every gap between two of them as a real contrast. But a gap
can be due to sampling randomness rather than genuine differences. The **combinatorial homogeneity test** of Le Roux,
Bienaise & Durand (2019, ch. 5) asks:

> Are the two groups **heterogeneous**?

It is a *permutation* test: keeping the group sizes fixed, it reallocates the
individuals to the groups in all possible ways (or many random ways) and asks
how often the reshuffled groups end up as far apart as the observed ones.

This is a different question from the one answered by the **typicality** test
(see `vignette("typicality")`). Typicality compares *one group with the whole
cloud* ("are the students unusual?"). Homogeneity compares *two groups with
each other* ("do the students and the retired differ?"). Two groups can both be
atypical of the main cloud yet be similar compared to one another.

For two groups the test reports:

* the **Mahalanobis distance `D`** between the two mean points (and `D²`);
* a **proportion of variance `η²`** -- the share of the cloud's variance that
  the split accounts for, the descriptive effect size (the book calls a
  difference *notable* when `η² ≥ 0.04`);
* a combinatorial **p-value**;
* a **compatibility region** for the deviation between the means.

## The cloud

We reuse the MCA of the `vignette("typicality")`: 2000 respondents of the French
"Histoire de vie" survey (`questionr::hdv2003`) and seven leisure practices,
with `sexe` and `occup` (activity status) as supplementary variables.

```{r mca}
library(GDAinference)
library(FactoMineR)
library(ggplot2)
data("hdv2003", package = "questionr")

active <- c("hard.rock", "lecture.bd", "peche.chasse",
            "cuisine", "bricol", "cinema", "sport")
d <- hdv2003[, c(active, "sexe", "occup", "age")]
mca <- MCA(d, quali.sup = 8:9, quanti.sup = 10, ncp = Inf, graph = FALSE)
```

On the first plane, **axis 1** is the *volume* of practices (few on the left,
many on the right; older respondents on the left) and **axis 2** opposes
*outdoor/manual* leisure (top) to *urban-cultural* leisure (bottom). The mean
point of each activity-status group -- the barycentre of its individuals, which
is what the test works on -- sits as follows:

```{r map, fig.alt = "Mean points of the activity-status groups on the first MCA plane."}
co <- get_coord(mca)[, 1:2]
occ <- mca$call$X$occup
bary <- aggregate(co, list(occup = occ), mean)
names(bary)[2:3] <- c("D1", "D2")

ggplot(bary, aes(D1, D2, label = occup)) +
  geom_hline(yintercept = 0, colour = "grey85") +
  geom_vline(xintercept = 0, colour = "grey85") +
  geom_point(colour = "#D95F02", size = 2.4) +
  geom_text(size = 3, fontface = "bold", vjust = -0.8) +
  scale_x_continuous(expand = expansion(mult = 0.25)) +
  coord_equal(clip = "off") + theme_minimal() +
  labs(x = "Axis 1 (volume of practices)",
       y = "Axis 2 (outdoor vs. urban-cultural)")
```

The students sit far in the "young / many / urban-cultural" corner and the
retired on the opposite side; the employed and the unemployed crowd near the
centre. Let us turn these impressions into tests.

## Do two groups differ? Students vs the retired

```{r students-retired}
sr <- homogeneity(mca, group = "occup",
                  groups = c("Etudiant, eleve", "Retraite"),
                  axes = 1:2, seed = 1, max_samples = 1e4)
sr
```

The mean points are far apart (`D ≈ 1.48`), the split is *notable*
(`η² ≈ 0.045 ≥ 0.04`) and **no** reshuffle out of 10,000 separates the two
groups as much -- `p ≈ 0`. Students and the retired are heterogeneous. The
`plot()` shows the inference directly, in the **space of deviations**:

```{r sr-plot, fig.alt = "Observed deviation between the two group means and its 95% compatibility region; the null point O lies far outside."}
plot(sr)
```

The orange point is the observed deviation $G_{c_2} - G_{c_1}$ (retired minus
students); it runs along **axis 1**, the volume of practices. The dashed
**95% compatibility region** around it is the set of deviations compatible with
the data. The null point **O** = "no difference" lies far outside it -- the
geometric face of `p < 0.05`. Note how *small* the region is: with hundreds of
individuals the deviation is pinned down tightly, so the difference is certified
with ease.

## Significance and size of differences

The homogeneity test, like the typicality test, separates *statistical
significance* from *the size of a difference*. Compare three pairs of groups:

```{r trichotomy}
pairs <- list(c("Exerce une profession", "Chomeur"),
              c("Au foyer",              "Retraite"),
              c("Etudiant, eleve",       "Retraite"))
do.call(rbind, lapply(pairs, function(p) {
  r <- homogeneity(mca, "occup", groups = p, axes = 1:2, seed = 1, max_samples = 1e4)
  data.frame(comparison = paste(p, collapse = " vs "),
             D = round(r$statistic, 2),
             eta2 = round(r$pv, 3),
             p_value = round(r$p_value, 4),
             notable = r$notable)
}))
```

Three contrasting situations, the whole point of the method:

* **Employed vs unemployed** -- `p ≈ 0.20`, `η² ≈ 0.001`. The two big central
  groups are **compatible**: their gap is small and could easily be due to the randomness of the sampling.
  On this plane they look alike. (Geometrically, **O** falls *inside* the
  compatibility region.)
* **Homemakers vs the retired** -- `p ≈ 0.002` but `η² ≈ 0.003`. The difference
  is **significant yet not notable**: with several hundred individuals the test
  detects a real but minute displacement (two groups that both sit on the
  low-volume left, barely apart). A small p-value is not equivalent to an important difference.
* **Students vs the retired** -- both notable and significant: a genuine,
  large heterogeneity.

## Partial or specific?

There are two ways to compare two groups (Le Roux et al. 2019, §5.3), and they
answer subtly different questions. So far we used the default, **partial**: the
two groups are situated *within the whole cloud* (the other respondents are kept
as a residual group and the Mahalanobis metric is the cloud's). Alternatively,
the **specific** comparison restricts the cloud to the two groups alone:

```{r specific}
homogeneity(mca, "occup", groups = c("Etudiant, eleve", "Retraite"),
            axes = 1:2, comparison = "specific", seed = 1, max_samples = 1e4)
```

The conclusion is unchanged (`p ≈ 0`), but the effect size jumps from
`η² ≈ 0.045` to `≈ 0.19`. The reason is instructive: *partial* `η²` is the share
of the **whole** cloud's variance explained by the split -- and students plus
retired are only a quarter of the 2000 respondents, so they can be far apart
(large `D`) yet account for little of the total variance. *Specific* `η²` is the
share within the **sub-cloud** of the two groups, where the same split explains
nearly a fifth of the variance. Use *partial* to weigh a contrast against the
whole analysis, *specific* to describe the two groups on their own terms.

When the two groups are the *only* groups in `group`, the comparison is the
"specific" one of Theorem 5.1, and the test then coincides exactly with the
combinatorial typicality test of one group against their union.

## A difference of a different kind: men and women

The direction of a difference matters as much as its size. Omitting `groups`
compares all the categories of the variable; for `sexe` that is simply men vs
women:

```{r sexe}
homogeneity(mca, group = "sexe", axes = 1:2, seed = 1, max_samples = 1e4)
```

Here `p ≈ 0` but `η² ≈ 0.03` (significant, not notable): a small but real gap.
Crucially it lies along **axis 2**, not axis 1 -- men lean toward
outdoor/manual leisure (`peche.chasse`, `bricol`), women toward urban-cultural
practices. Men and women do *similar amounts* (axis 1) of *different kinds*
(axis 2) of leisure. The occupation contrasts above, by contrast, were
differences of *volume* along axis 1. Homogeneity tells you not just *whether*
two groups differ, but *in which direction*.

## All groups at once: the omnibus

To ask whether the activity-status groups are heterogeneous *as a whole*, omit
`groups` so that all seven categories are compared.
With more than two groups the statistic is the between-group Mahalanobis
variance `V_M` (an omnibus test; there is no single deviation, hence no region):

```{r omnibus}
homogeneity(mca, group = "occup", axes = 1:2, seed = 1, max_samples = 1e4)
```

`η² ≈ 0.11` of the cloud's variance lies between the activity-status groups, and
`p ≈ 0`: the groups are strongly heterogeneous overall. This is the homogeneity
counterpart of an omnibus "are these groups different?" before drilling into the
specific pairwise contrasts above.

## Typicality or homogeneity?

The two tests of the package answer neighbouring but distinct questions, and the
cloud reading tells you which you need:

* [typicality_comb()] -- *one group vs the whole cloud*. "Are the students
  atypical of the respondents?" (`vignette("typicality")`).
* `homogeneity()` -- *group vs group*. "Do the students and the retired differ
  from each other?", or "are the activity-status groups heterogeneous?"

They can disagree in illuminating ways: two groups may both be atypical of the
cloud yet not heterogeneous from each other, or both be
typical yet still differ; and a single large group can be significantly atypical
while its difference from a neighbour is negligible.

## Practical notes

* **Exact vs. Monte Carlo.** When the number of reshuffles is small the test is
  exact; on a cloud of 2000 it draws `max_samples` of them (default `1e5`; set
  `seed` for reproducibility). To keep this vignette quick we capped them at
  `max_samples = 1e4`; with such tiny p-values, drawing more only sharpens an
  already-clear conclusion.
* **Which axes?** The test runs in the subspace you pass through `axes`. We read
  it on the first plane because that is what we interpreted; pass `axes = NULL`
  (with `ncp = Inf` in the MCA) to test in the full cloud.
* **One dimension.** On a single axis the test becomes a signed, one-sided
  comparison of the two group means, with an exact compatibility *interval* --
  useful to test a contrast on one interpreted axis. The direction is taken
  from the observed difference, so for a *non-directional* claim compare the
  one-sided p-value to $\alpha/2$ (this matches the interval, which excludes
  zero exactly when $p < \alpha/2$).
* **Other GDA packages.** The same calls work on a GDAtools `speMCA` / `csMCA`
  or an ade4 `dudi.pca` / `dudi.acm`; pass that object, and the grouping
  variable as a vector when it is not stored in the object.

## Reference

Le Roux, B., Bienaise, S. & Durand, J.-L. (2019).
*Combinatorial Inference in Geometric Data Analysis*. Chapman & Hall/CRC.
