---
title: "Combinatorial typicality testing on an MCA: a tutorial"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Combinatorial typicality testing on an MCA: a tutorial}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
needed <- c("questionr", "FactoMineR", "ggplot2")
ok <- all(vapply(needed, requireNamespace, logical(1), quietly = TRUE))
knitr::opts_chunk$set(
  collapse = TRUE, comment = "#>", eval = ok,
  fig.width = 7, fig.height = 5.2, fig.align = "center"
)
```

## Why inference in GDA?

Geometric Data Analysis — MCA, PCA, CA and their variants — represents the
individuals of a data table as a *cloud of points*: individuals with similar
profiles are close to one another, dissimilar ones far apart. The principal
axes of the cloud summarise the main oppositions in the data, and
supplementary variables (sex, age, occupation, …) can be projected onto the
cloud to locate groups of individuals within it. In practice the principal
axes are often read as underlying dimensions structuring the data — much as
one would read latent variables — although GDA itself stays deliberately
closer to the data: the cloud, not a hypothesised construct, is the object
under study.

The traditional way of interpreting such an analysis is descriptive:

1. give meaning to the axes, from the categories that contribute most to them;
2. project the supplementary categories and read the *oppositions between
   groups* — "students lie on the side of numerous, urban-cultural practices;
   the retired on the opposite side";
3. sometimes, back this reading with the axis-by-axis indicators offered by
   some GDA software (test values based on a normal/chi-squared
   approximation).

This workflow leaves one question unanswered. A group's mean point *always*
deviates somewhat from the centre of the cloud, so the deviation one sees
proves nothing by itself: could a deviation of that size arise for a randomly
drawn subset of the same size — sampling variability, in classical terms — or
is the group genuinely located elsewhere than the rest of the cloud? The eye
cannot decide, and the axis-by-axis indicators answer only axis by axis, and
only approximately, whereas the deviation is a genuinely multidimensional
quantity: a group can be unremarkable on each axis taken separately yet occupy
a remarkable position in the plane.

The **combinatorial typicality test** of Le Roux, Bienaise & Durand (2019),
which `GDAinference` implements, answers that question exactly. It is a
*permutation* test, so it makes no assumption about the distribution of the
data: it compares the group's observed deviation with the deviation of *all
possible* groups of the same size drawn from the cloud. It returns two things:

* a **p-value** — the proportion of random groups that deviate at least as
  much as the observed one (small p ⇒ the group is *atypical*: it cannot be
  assimilated to a random subset of the cloud);
* a **compatibility region** — an ellipse showing where the group mean could
  plausibly lie.

Throughout, the test statistic is the **Mahalanobis distance `D`** between the
group mean and the cloud centre. As a rule of thumb the deviation is called
*notable* when `D ≥ 0.4`. Keep that number in mind — we will see that *notable*
and *significant* are not the same thing.

We will work end to end on a real survey.

## The data and the MCA

We use `hdv2003` from **questionr** (2000 respondents to the French
"Histoire de vie" survey, 2003) and seven yes/no leisure practices. The
variables carry French names and labels, so we first relabel them in English:

```{r load}
library(GDAinference)
library(FactoMineR)
data("hdv2003", package = "questionr")

d <- hdv2003[, c("hard.rock", "lecture.bd", "peche.chasse", "cuisine",
                 "bricol", "cinema", "sport", "sexe", "occup", "age")]
names(d) <- c("hard.rock", "comics", "fishing.hunting", "cooking",
              "DIY", "cinema", "sport", "sex", "occupation", "age")
for (v in names(d)[1:7]) levels(d[[v]]) <- c("No", "Yes")
levels(d$sex) <- c("Man", "Woman")
levels(d$occupation) <- c("Employed", "Unemployed", "Student", "Retired",
                          "Retired from business", "Homemaker",
                          "Other inactive")
```

We build a standard MCA of the seven practices with **FactoMineR**, adding
`sex` and `occupation` as supplementary categorical variables and `age` as a
supplementary numeric variable, so they do not shape the cloud but can be
located on it:

```{r mca}
mca <- MCA(d, quali.sup = 8:9, quanti.sup = 10, ncp = Inf, graph = FALSE)
round(mca$eig[1:3, ], 2)
```

The first plane (axes 1–2) carries the main structure, so we will read the
typicality tests on it. (MCA's raw inertia rates are notoriously pessimistic;
what matters here is the *geometry* of the plane, not the percentages.)

## Reading the cloud

Before testing anything, we must understand what the axes *mean* — otherwise a
p-value is just a number. The map below shows the active categories (red) and
the supplementary categories (green), among which the occupation groups:

```{r map, fig.alt = "Map of the first MCA plane: active categories and supplementary categories."}
plot(mca, invisible = "ind", cex = 0.8)
```

Two clear oppositions emerge:

* **Axis 1 — the *volume* of practices.** On the right sit the "Yes"
  categories (`hard.rock_Yes`, `comics_Yes`, `sport_Yes`, `cinema_Yes`); on
  the left, the "No" categories. The right means *does many things*, the left
  *does few*. The supplementary variable `age` correlates −0.44 with this
  axis: **older respondents are on the left** (fewer practices), younger on
  the right.
* **Axis 2 — the *kind* of practice.** At the top, outdoor/manual leisure
  (`fishing.hunting_Yes` stands out, with `DIY_Yes`); at the bottom,
  urban-cultural leisure (`comics_Yes`, `cinema_Yes`, `hard.rock_Yes`).

The occupation points already tell a story: **students** sit far to the lower
right (young, many and urban-cultural practices), **retired** people to the
left (older, fewer practices), while the **employed** and the **unemployed**
sit close to the centre. Typicality testing turns this visual impression into
a formal statement.

## A first test: are students atypical?

```{r students}
students <- typicality_comb(mca, group = "occupation", level = "Student",
                            axes = 1:2, seed = 1)
students
```

The Mahalanobis distance is `D ≈ 0.97` — well above the 0.4 *notable* limit —
and the p-value is ≈ 0: among all groups of 94 individuals one could draw from
the cloud, essentially none deviates as much as the students do. Students are
**atypical**, and the map tells us *how*: they are pulled toward the
"young / many / urban-cultural" corner.

The `plot()` method makes this geometric:

```{r students-plot, fig.alt = "Compatibility ellipse for the students' mean point."}
plot(students)
```

The dashed ellipse is the **95% compatibility region** for the students' mean
point **C** (orange). The cloud centre **G** (black) lies far outside it, which
is the geometric face of `p < 0.05`. Notice how *small* the ellipse is: with
2000 individuals the mean is pinned down very precisely, so even moderate
deviations become detectable. More data ⇒ smaller region ⇒ easier to certify
that a deviation is not one that a random subset could produce.

## *Significant* is not *notable*

Let us now test every occupation group at once, on the plane:

```{r byvar}
typicality_byvar(mca, "occupation", axes = 1:2, seed = 1)
```

Read this table carefully — it contains the whole philosophy of the method:

* **Unemployed**: `D ≈ 0.13`, p ≈ 0.31. The deviation is tiny *and* not
  significant: the unemployed are **compatible** with the cloud as a whole —
  on this plane they look like everyone else (they sit near the centre).
* **Employed** and **Homemaker**: `p ≈ 0`, yet `D < 0.4`. These deviations are
  **statistically significant but descriptively small**. With n = 1049
  employed respondents, the test detects a real but minute displacement. This
  is the large-sample caveat that the *notable* threshold is designed to
  catch: **do not mistake a significant p-value for an important effect.**
  Always read `D` alongside `p`.
* **Student, Retired, Retired from business, Other inactive**: both `notable`
  (`D ≥ 0.4`) **and** significant. These are the genuinely atypical groups —
  and, reading the map, they are atypical in *opposite* directions along
  axis 1 (students young/active, retired older/less active).

This three-way reading — *compatible* / *significant-but-small* / *notable and
significant* — is exactly what combinatorial inference is meant to support.

## The v-test on a supplementary category: what's the difference?

Readers used to FactoMineR will know another indicator attached to a
supplementary category: the **v-test** (*valeur-test*). It answers the *same*
underlying question — could this category's mean point be that of a random
subset of the individuals? — so it is worth seeing exactly how the two relate.
Here are the v-tests of the occupation groups on the first two axes:

```{r vtest}
round(mca$quali.sup$v.test[levels(d$occupation), 1:2], 2)
```

The v-test is, in effect, the approximate, one-axis ancestor of the typicality
test (the book discusses it in §3.2.6). Three things separate them:

* **Axis by axis vs. multidimensional.** A v-test is computed per axis, so a
  group gets one number per dimension (and testing several axes raises an
  implicit multiple-comparison problem). The typicality test works on the
  deviation *in the plane* (or any subspace), with the Mahalanobis metric of
  the cloud: one test for the position the eye actually reads on the map.
* **Approximate vs. exact.** The v-test relies on a normal approximation of
  the mean of a random subset (hypergeometric sampling); the typicality test
  computes the actual combinatorial distribution, exhaustively or by Monte
  Carlo.
* **Significance vs. magnitude, again.** The magnitude of a v-test conflates
  the size of the deviation with the size of the group — it grows like
  $\sqrt{n_c}$ — and it comes with no effect size and no compatibility
  region. The typicality test separates the two (`D` alongside `p`) and
  returns the region.

The table above makes the third point concrete. Ranked by v-test on axis 1,
**Employed** (13.1) beats **Student** (7.4) — it looks like the most
strikingly positioned group of all. Yet the typicality table showed Employed
at `D = 0.28` (significant but *not* notable) against Student's `D = 0.97`:
the employed simply outnumber everyone (n = 1049), so even their minute
displacement produces a huge v-test. Conversely, **Unemployed** (v-tests 1.4
and 0.6, both below the conventional |2| threshold) agrees with the
typicality verdict `p ≈ 0.31`: the two approaches converge where the
approximation is comfortable, and the typicality test adds the exactness, the
multidimensional reading and the effect size where it matters.

## Practical notes

* **Which axes?** The test runs in the subspace you pass through `axes`. Here we
  used the first plane (`axes = 1:2`) because that is what we interpreted. To
  test the deviation in the *full* cloud, keep all axes (`ncp = Inf` in the MCA)
  and call with `axes = NULL`.
* **Naming vs. passing a variable.** Because `occupation` is stored in the MCA
  object, we referred to it by name (`"occupation"`). You can equally pass the
  vector (`group = d$occupation`), which is the way to go for objects that do
  not store the variable (e.g. GDAtools `speMCA`).
* **Exact vs. Monte Carlo.** With small clouds the test enumerates *all* subsets
  (exact); with large ones like this it draws `max_samples` of them. Set `seed`
  for reproducible Monte Carlo p-values.
* **Other GDA packages.** The very same calls work on a GDAtools `speMCA` /
  `csMCA` or an ade4 `dudi.pca` / `dudi.acm` — just pass that object instead of
  the FactoMineR one.

## Reference

Le Roux, B., Bienaise, S. & Durand, J.-L. (2019).
*Combinatorial Inference in Geometric Data Analysis*. Chapman & Hall/CRC.
