---
title: "Combinatorial typicality test"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Combinatorial typicality test}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

## The typicality problem

In Geometric Data Analysis (PCA, MCA, CA, ...) we often observe that the mean
point of a group of individuals lies far from the centre of the cloud, and we
want to know whether this deviation is a *genuine* effect or could plausibly be
due to chance. The **combinatorial typicality test** of
Le Roux, Bienaise and Durand (2019, chapter 3) answers this question without any
assumption about the distribution of the data.

It uses as test statistic the squared **Mahalanobis distance** $D^2$ between the
group mean point and the cloud centre, computed with respect to the covariance
structure of the reference cloud. Its reference distribution is obtained
*combinatorially*: by considering every subset of $n_c$ points that can be drawn
from the $n$ reference points (exhaustively when feasible, by Monte Carlo
otherwise). The p-value is the proportion of subsets whose statistic is at least
as large as the observed one.

```{r setup}
library(GDAinference)
```

## The "Target" example

We reproduce the worked example of the book. The reference cloud `Target` is a
set of 10 points in a plane; the group `Target_group` is a set of 4 points whose
mean we compare with the cloud centre.

```{r data}
Target
Target_group
```

## Running the test

```{r run}
res <- typicality_comb(Target, Target_group)
res
```

The Mahalanobis distance is `D = 0.964`, descriptively *notable* (the rule of
thumb is a limit of 0.4). The test enumerates the
$\binom{10}{4} = 210$ possible subsets exhaustively and finds 9 of them at least
as extreme as the observed deviation, hence an exact p-value of
$9/210 = 0.043$ -- matching the book.

Every quantity is available programmatically:

```{r components}
res$statistic              # Mahalanobis distance D
res$p_value                # combinatorial p-value
res$n_sup                  # subsets at least as extreme ...
res$n_perm                 # ... out of this many
res$compatibility$d_alpha  # scale of the 95% compatibility ellipsoid
```

## The compatibility region

The `plot()` method shows the reference cloud, the centre **G**, the group mean
**C**, and the 95% compatibility region -- the principal $d_\alpha$-ellipse of
the reference cloud centred on **C** (here $d_\alpha = 0.957$). Point **G** lies
just outside it, the geometric counterpart of $p < 0.05$.

```{r plot, eval = requireNamespace("ggplot2", quietly = TRUE), fig.width = 6, fig.height = 5, fig.alt = "Target reference cloud with the 95% compatibility ellipse centred on the group mean."}
plot(res)
```

## Specifying the group as a subset

When the group is a subset of the reference individuals, you can pass a logical
vector, or row indices / names, instead of a separate coordinate table:

```{r subset}
typicality_comb(Target, group = c("i1", "i3", "i5"))$p_value
```

## Working from a GDA result

`typicality_comb()` accepts the result object of the common GDA packages
directly: the principal coordinates are extracted automatically with
`get_coord()`. For instance, with **FactoMineR** (the same pattern applies to
**GDAtools** `speMCA()`/`csMCA()` and **ade4** `dudi.*()` objects):

```{r gda, eval = FALSE}
library(FactoMineR)
mca <- MCA(my_data, graph = FALSE, ncp = Inf)        # keep all axes
typicality_comb(mca, group = my_data$category == "workers", axes = 1:3)
```

Note that the test is carried out in the subspace spanned by the requested
`axes`; to obtain the genuine full-cloud test, retain all principal axes in the
upstream analysis (`ncp = Inf`).

## Reference

Le Roux, B., Bienaise, S. & Durand, J.-L. (2019).
*Combinatorial Inference in Geometric Data Analysis*. Chapman & Hall/CRC.
