---
title: "Combinatorial typicality testing on an MCA: a tutorial"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Combinatorial typicality testing on an MCA: a tutorial}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
needed <- c("questionr", "FactoMineR", "ggplot2")
ok <- all(vapply(needed, requireNamespace, logical(1), quietly = TRUE))
knitr::opts_chunk$set(
  collapse = TRUE, comment = "#>", eval = ok,
  fig.width = 7, fig.height = 5.2, fig.align = "center"
)
```

## What this tutorial is about

Geometric Data Analysis (MCA, PCA, CA, …) builds a *cloud of individuals* and
lets us describe oppositions between groups. But description alone leaves a
question open: when a group's mean point sits away from the centre of the cloud,
**is that a genuine effect, or could it just be the luck of the draw?**

`GDAinference` answers this with the **combinatorial typicality test** of
Le Roux, Bienaise & Durand (2019). It is a *permutation* test, so it makes no
assumption about the distribution of the data: it compares the group's observed
position to what would happen for *all possible* groups of the same size drawn
from the cloud. It returns two things:

* a **p-value** — the proportion of random groups that deviate at least as much
  as the observed one (small p ⇒ the group is *atypical*);
* a **compatibility region** — an ellipse showing where the group mean could
  plausibly lie.

Throughout, the test statistic is the **Mahalanobis distance `D`** between the
group mean and the cloud centre. As a rule of thumb the deviation is called
*notable* when `D ≥ 0.4`. Keep that number in mind — we will see that *notable*
and *significant* are not the same thing.

We will work end to end on a real survey.

## The data and the MCA

We use `hdv2003` from **questionr** (2000 respondents to the French
"Histoire de vie" survey) and seven yes/no cultural and leisure practices. We
build a standard MCA with **FactoMineR**, adding `sexe` and `occup` (activity
status) as supplementary variables and `age` as a supplementary numeric
variable, so they do not shape the cloud but can be located on it.

```{r load}
library(GDAinference)
library(FactoMineR)
library(ggplot2)
data("hdv2003", package = "questionr")

active <- c("hard.rock", "lecture.bd", "peche.chasse",
            "cuisine", "bricol", "cinema", "sport")
d <- hdv2003[, c(active, "sexe", "occup", "age")]

mca <- MCA(d, quali.sup = 8:9, quanti.sup = 10, ncp = Inf, graph = FALSE)
round(mca$eig[1:3, ], 2)
```

The first plane (axes 1–2) carries the main structure, so we will read the
typicality tests on it. (MCA's raw inertia rates are notoriously pessimistic;
what matters here is the *geometry* of the plane, not the percentages.)

## Reading the cloud

Before testing anything, we must understand what the axes *mean* — otherwise a
p-value is just a number. The map below shows the active categories (grey) and
the activity-status groups (orange).

```{r map, fig.alt = "Map of the first MCA plane: active categories and occupation groups."}
cats <- function(coord, type) {
  df <- as.data.frame(coord[, 1:2]); names(df) <- c("D1", "D2")
  data.frame(label = rownames(df), D1 = df$D1, D2 = df$D2, type = type)
}
map <- rbind(
  cats(mca$var$coord, "leisure category"),
  cats(mca$quali.sup$coord[levels(hdv2003$occup), ], "occupation")
)

ggplot(map, aes(D1, D2, colour = type, label = label)) +
  geom_hline(yintercept = 0, colour = "grey85") +
  geom_vline(xintercept = 0, colour = "grey85") +
  geom_point(size = 1.6) +
  # thin out the leisure labels, but always show the (few) occupation labels
  geom_text(data = subset(map, type == "leisure category"),
            size = 2.8, vjust = -0.7, check_overlap = TRUE, show.legend = FALSE) +
  geom_text(data = subset(map, type == "occupation"),
            size = 3, fontface = "bold", vjust = -0.8, show.legend = FALSE) +
  scale_colour_manual(values = c("leisure category" = "grey45",
                                 "occupation" = "#D95F02")) +
  scale_x_continuous(expand = expansion(mult = 0.18)) +
  scale_y_continuous(expand = expansion(mult = 0.08)) +
  coord_equal(clip = "off") + theme_minimal() +
  labs(x = "Axis 1 (volume of practices)",
       y = "Axis 2 (outdoor vs. urban-cultural)", colour = NULL)
```

Two clear oppositions emerge:

* **Axis 1 — the *volume* of practices.** On the right sit the "Oui" categories
  (`hard.rock`, `lecture.bd`, `sport`, `cinema`); on the left, the "Non"
  categories. The right means *does many things*, the left *does few*. The
  supplementary variable `age` correlates −0.44 with this axis: **older
  respondents are on the left** (fewer practices), younger on the right.
* **Axis 2 — the *kind* of practice.** At the top, outdoor/manual leisure
  (`peche.chasse` stands out, with `bricol`); at the bottom, urban-cultural
  leisure (`lecture.bd`, `cinema`, `hard.rock`).

The orange occupation points already tell a story: **students** sit far to the
lower right (young, many and urban-cultural practices), **retired** people to
the left (older, fewer practices), while the **employed** and the **unemployed**
sit close to the centre. Typicality testing turns this visual impression into a
formal statement.

## A first test: are students atypical?

```{r students}
students <- typicality_comb(mca, group = "occup", level = "Etudiant, eleve",
                            axes = 1:2, seed = 1)
students
```

The Mahalanobis distance is `D ≈ 0.97` — well above the 0.4 *notable* limit —
and the p-value is ≈ 0: among all groups of 94 individuals one could draw from
the cloud, essentially none deviates as much as the students do. Students are
**atypical**, and the map tells us *how*: they are pulled toward the
"young / many / urban-cultural" corner.

The `plot()` method makes this geometric:

```{r students-plot, fig.alt = "Compatibility ellipse for the students' mean point."}
plot(students)
```

The dashed ellipse is the **95% compatibility region** for the students' mean
point **C** (orange). The cloud centre **G** (black) lies far outside it, which
is the geometric face of `p < 0.05`. Notice how *small* the ellipse is: with
2000 individuals the mean is pinned down very precisely, so even moderate
deviations become detectable. More data ⇒ smaller region ⇒ easier to certify a
genuine effect.

## The key lesson: *significant* is not *notable*

Let us now test every activity-status group at once, on the plane:

```{r byvar}
typicality_byvar(mca, "occup", axes = 1:2, seed = 1)
```

Read this table carefully — it contains the whole philosophy of the method:

* **Chômeur** (unemployed): `D ≈ 0.13`, p ≈ 0.31. The deviation is tiny *and*
  not significant: the unemployed are **compatible** with the cloud as a whole —
  on this plane they look like everyone else (they sit near the centre).
* **Exerce une profession** (employed) and **Au foyer** (homemakers):
  `p ≈ 0`, yet `D < 0.4`. These deviations are **statistically significant but
  descriptively small**. With n = 1049 employed respondents, the test detects a
  real but minute displacement. This is the large-sample caveat that the
  *notable* threshold is designed to catch: **do not mistake a significant
  p-value for an important effect.** Always read `D` alongside `p`.
* **Students, retired, "retired from business", other inactive**: both
  `notable` (`D ≥ 0.4`) **and** significant. These are the genuinely atypical
  groups — and, reading the map, they are atypical in *opposite* directions
  along axis 1 (students young/active, retired older/less active).

This three-way reading — *compatible* / *significant-but-small* / *notable and
significant* — is exactly what combinatorial inference is meant to support.

## Practical notes

* **Which axes?** The test runs in the subspace you pass through `axes`. Here we
  used the first plane (`axes = 1:2`) because that is what we interpreted. To
  test the deviation in the *full* cloud, keep all axes (`ncp = Inf` in the MCA)
  and call with `axes = NULL`.
* **Naming vs. passing a variable.** Because `occup` is stored in the MCA
  object, we referred to it by name (`"occup"`). You can equally pass the vector
  (`group = hdv2003$occup`), which is the way to go for objects that do not store
  the variable (e.g. GDAtools `speMCA`).
* **Exact vs. Monte Carlo.** With small clouds the test enumerates *all* subsets
  (exact); with large ones like this it draws `max_samples` of them. Set `seed`
  for reproducible Monte Carlo p-values.
* **Other GDA packages.** The very same calls work on a GDAtools `speMCA` /
  `csMCA` or an ade4 `dudi.pca` / `dudi.acm` — just pass that object instead of
  the FactoMineR one.

## Reference

Le Roux, B., Bienaise, S. & Durand, J.-L. (2019).
*Combinatorial Inference in Geometric Data Analysis*. Chapman & Hall/CRC.
